Interactive Visualization of Hierarchically Structured Data

### 2016-12-04 treelapse.Rmd
Revisions to problem motivation

We have implemented our visualizations in d3, but encapsulated in an R
package, to facilitate rapid turnover from data preparation and
modeling to interactive exploration, and vice versa. We hope this
package encourages data analysts to work at the border between data
modeling and visualization, and more generally empowers a wider
audience to apply less widely known, but powerful, visualization
ideas.

The paper is organized as follows. First, we describe our microbiome
application, and then abstract away the essential analysis tasks. This
abstraction suggests applications to standard problems outside of the
microbiome, some of which we articulate here. Next, we review the underlying
visualization principles behind our work, and provide references to the
software systems that originally implemented them. We then connect these
principles to analysis tasks we identified earlier, describing in detail the
visualization methods we have implemented in treelapse. We close with several
case studies using publicly available data.

@@

Microbiomics is the study of the communities of microbes living in given
environments, for example, ocean water or the human gut. Progress in the field
has been rapidly accelerated by the advent of genomic technologies, which
enable detailed quantification of microbial community structure, and as the
relationship between the microbiome and human health has become more and more
apparent, research into the human microbiome has garnered widespread
attention. Being concerned with both microbial community structure and human
health, the field exists at the border between ecology and medicine;
consequently, papers in the area often apply a blend of exploratory data
analysis and formal statistical inference ideas.

The two essential microbiome analysis problems that motivated our work are the
tree structured differential abundance and microbial dynamics problems. In the
differential abundance problem, we attempt to compare the abundances of
individual microbes across experimental conditions -- for example, treatment
vs. control or healthy vs. diseased. This is the microbiome analog of
differential expression analysis in genomics. We prepend the description
"tree-structured" because, in practice, researchers generate interpretations
about middling taxonomic orders of microbes -- it is more interesting to
discover novel behavior taxonomic levels intermediate between high-order phyla
and low-level sequence variants. Hence, we frame the tree-structured
differential abundance problem as the question of identifying the largest
taxonomic subtree whose associated microbes are differentially abundant.

In the tree structured microbial dynamics problem, the goal is to describe
changes in microbial abundances in an environment over time. As in the
differential abundance problem, it is useful if these descriptions can be given
at the highest subtree at which the pattern appears. Specific questions of
interest often have an ecological flavor. For example, researchers are often
interested in understanding how microbial populations respond to sudden or
gradual environmental changes or how microbes fill, drop out from, or compete
for certain environmental niches. Medically, these questions are important for
illuminating the impact of antibiotic time courses or the potential for direct
manipulation, for example.

>>>>

We have implemented our visualizations in d3, but encapsulated in an R
package, called treelapse, to facilitate rapid turnover from data
preparation and modeling to interactive exploration, and vice
versa. We hope this package encourages data analysts to work at the
border between data modeling and visualization, and more generally
empowers a wider audience to apply less widely known, but powerful,
visualization ideas.

The paper is organized as follows. First, we describe our motivating
application to the microbiome and the associated generic analysis
tasks. Next, we review the underlying visualization principles behind
our contributions. Then we then connect these principles to analysis
tasks we identified earlier, describing in detail the visualization
methods we have implemented in treelapse. We close with several case
studies using publicly available data across both microbiome and
non-microbiome related applications.

@@

A microbiome is a community of microbes living in given environments,
for example, ocean water or the human gut [@human2012structure,
@human2012structure, @cho2012human]. Progress in the field has been
rapidly accelerated by the advent of genomic technologies, which
enable detailed quantification of microbial ecological structure and
its influence in human and environmental health. Being concerned with
both microbial community structure and human health, the field exists
at the border between ecology and medicine; consequently, papers in
the area often apply a blend of exploratory data analysis and formal
statistical inference.

The two essential microbiome analysis problems that motivated our work
are the tree-structured differential abundance and differential
dynamics problems. In the differential abundance problem, we attempt
to compare the abundances of individual microbes across experimental
conditions -- for example, treatment vs. control or healthy
vs. diseased. This is the microbiome analog of differential expression
analysis in genomics [@anders2010differential]. We prepend the
description "tree-structured" because, in practice, researchers
generate interpretations about intermediate taxonomic orders of
microbes -- it is more interesting to discover novel behavior
taxonomic levels between high-order phyla and low-level
species. Hence, we frame the tree-structured differential abundance
problem as the question of identifying the largest taxonomic subtree
whose associated microbes are differentially abundant.

In the tree structured microbial dynamics problem, the goal is to
describe changes in microbial abundances in an environment over
time. As in the differential abundance problem, it is useful if these
descriptions can be given at the highest subtree at which the pattern
appears. Specific questions of interest often have an ecological
flavor. For example, researchers are often interested in understanding
how microbial populations respond to sudden or gradual environmental
changes or how microbes fill, drop out from, or compete for
environmental niches. Medically, these questions are important for
illuminating the impact of antibiotic time courses or diet changes,
for example.
